# Reading Data into R

**Chapter 4, *Last updated: Jan 1, 2023.***

Dataframes and Tibbles are frequently employed data structures in R for storing and manipulating data. They facilitate the organization, exploration, and analysis of data.

## Dataframes

1.  A dataframe is a two-dimensional table-like data structure in R that stores data in rows and columns, with distinct data types for each column.

2.  Similar to a spreadsheet or a SQL table, it is one of the most frequently employed data structures in R. Each column in a data.frame is a constant-length vector, and each row represents an observation or case.

3.  dataframe objects can be created in R using the `data.frame()` function or by importing data from external sources such as CSV files, Excel spreadsheets, or databases.

4.  R has many useful built-in methods and functions for manipulating and summarizing data stored in dataframes, including subsetting, merging, filtering, and aggregation. \[1\]

### Creating a dataframe using raw data

5.  The following code generates a data.frame named `df` containing three columns - `names`, `ages`, and `heights`, and four rows of data for each individual.

```{r}
# Create input data as vectors
names <- c("Ashok", "Bullu", "Charu", "Divya")
ages <- c(72, 49, 46, 42)
heights <- c(170, 167, 160, 166)

# Combine input data into a data.frame
people <- data.frame(Name = names, Age = ages, Height = heights)

# Print the resulting dataframe
print(people)
```

## Reading Inbuilt datasets in R

1.  R contains a number of built-in datasets that can be accessed without downloading or integrating from external sources. Here are some of the most frequently used built-in datasets in R:

-   `women`: This dataset includes the heights and weights of a sample of 15,000 women.

-   `mtcars`: This dataset contains information on 32 distinct automobile models, including the number of cylinders, engine displacement, horsepower, and weight.

-   `diamonds`: This dataset includes the prices and characteristics of approximately 54,000 diamonds, including carat weight, cut, color, and clarity.

-   `iris`: This data set measures the sepal length, sepal width, petal length, and petal breadth of 150 iris flowers from three distinct species.

### The `women` dataset

As an illustration, consider the `women` dataset inbuilt in R, which contains information about the heights and weights of women. \[2\] It has just two variables:

1.  `height`: Height of each woman in inches

2.  `weight`: Weight of each woman in pounds

3.  The `data()` function is used to import any inbuilt dataset into R. The `data(women)` command in R loads the `women` dataset

```{r}
data(women)
```

4.  The `str()` function gives the dimensions and data types and also previews the data.

```{r}
str(women)
```

5.  The `summary()` function gives some summary statistics.

```{r}
summary(women)
```

### The `mtcars` dataset

The `mtcars` dataset inbuilt in R comprises data on the fuel consumption and other characteristics of 32 different automobile models. \[2\]

Here is a concise description of the 11 `mtcars` data columns:

1.  `mpg`: Miles per gallon (fuel efficiency)

2.  `cyl`: Number of cylinders

3.  `disp`: Displacement of the engine (in cubic inches)

4.  `hp`: gross horsepower

5.  `drat`: Back axle ratio wt: Weight (in thousands of pounds)

6.  `wt`: Weight (in thousands of pounds)

7.  `qsec`: 1/4 mile speed (in seconds)

8.  `vs`: Type of engine (0 = V-shaped, 1 = straight)

9.  `am`: Type of transmission (0 for automatic, 1 for manual)

10. `gear`: the number of forward gears

11. `carb`: the number of carburetors

```{r}
data(mtcars)
str(mtcars)
```

## Reading different file formats into a dataframe

1.  We examine how to read data into a dataframe in R when the original data is stored in prominent file formats such as CSV, Excel, and Google Sheets. \[3\]

2.  Before learning how to accomplish this, it is necessary to comprehend how to configure the Working Directory in R.

### Working Directory

1.  The working directory is the location where R searches for and saves files by default.

2.  By default, when we execute a script or import data into R, R will search the working directory for files.

3.  Using R's `getwd()` function, we can examine our current working directory:

```{r}
getwd()
```

4.  We are running R in the Cloud and hence we are seeing that the working directory is specified as `/cloud/project/DataAnalyticsBook101`. If we are doing R programming on a local computer, and if our working directory is the Desktop, then we may see a different response such as `C:/Users/YourUserName/Desktop`.

5.  Using R's `setwd()` function, we can change our current working directory. For example, the following code will set our working directory to the Desktop:

```{r eval=FALSE, echo=TRUE}
#setwd("C:/Users/YourUserName/Desktop")
```

6.  We should choose an easily-remembered and accessible working directory to store our R scripts and data files. Additionally, we should avoid using spaces, special characters, and non-ASCII characters in file paths, as these can cause file handling issues in R. \[3\]

### Reading a CSV file into a dataframe

1.  CSV is the abbreviation for "Comma-Separated Values." A CSV file is a plain text file that stores structured tabular data.

2.  Each entry in a CSV file represents a record, whereas each column represents a field. The elements in each record are separated by commas (hence the name Comma-Separated Values), semicolons, or tabs.

3.  Before proceeding ahead, it is imperative that the file that we wish to read is located in the Working Directory.

4.  Suppose we wish to import a CSV file named `mtcars.csv`, located in the Working Directory. We can use the `read.csv()` function, illustrated as follows.

```{r eval=FALSE}
df_csv <- read.csv("mtcars.csv")
```

4.  In this example, the `read.csv()` function reads the mtcars.csv file into a data frame named `df_csv`.

5.  If the file is not in the current working directory, the complete file path must be specified in the `read.csv()` function argument; otherwise, an error will occur. \[3\]

### Reading an Excel (xlsx) file into a dataframe

1.  Suppose we wish to import a Microsoft Excel file named `mtcars.xlsx`, located in the Working Directory.

2.  We can use the `read_excel` function in the R package `readxl`, illustrated as follows. \[3\]

```{r eval=FALSE}
library(readxl)
df_xlsx <- read_excel("mtcars.xlsx")
```

### Reading a Google Sheet into a dataframe

1.  Google Sheets is a ubiquitous cloud-based spreadsheet application developed by Google. It is a web-based application that enables collaborative online creation and modification of spreadsheets. \[4\]

2.  We can import data from a Google Sheet into a R dataframe, as follows.

-   Consider a Google Sheet whose preferences have been set such that anyone can view it using its URL. If this is not done, then some authentication would become necessary.

-   Every Google Sheet is characterized by a unique Sheet ID, embedded within the URL. For example, consider a Google Sheet containing some financial data concerning S&P500 index shares.

-   Suppose the Sheet ID is: `11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM`

-   We can use the function `gsheet2tbl` in package `gsheet` to read the Google Sheet into a dataframe, as demonstrated in the following code.

```{r, warning=FALSE}
# Read S&P500 stock data present in a Google Sheet.
library(gsheet)

prefix <- "https://docs.google.com/spreadsheets/d/"
#sheetID <- "11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM"
sheetID <- "1F5KvFATcehrdJuGjYVqppNYC9hEKSww9rXYHCk2g6OA"

# Form the URL to connect to
url500 <- paste(prefix, sheetID) 

# Read the Google Sheet located at the URL into a dataframe called sp500
sp500 <- gsheet2tbl(url500)
```

-   The first line imports the `gsheet` package required to access Google Sheets into R.

-   The following three lines define URL variables for Google Sheets. The `prefix` variable contains the base URL for accessing Google Sheets, the `sheetID` variable contains the ID of the desired Google Sheet.

-   The `paste()` function is used to combine the `prefix`, `sheetID` variables into a complete URL for accessing the Google Sheet.

-   The `gsheet2tbl()` function from the `gsheet` package is then used to read the specified Google Sheet into a dataframe called `sp500`, which can then be analyzed further in R. \[4\]

### Joining or Merging two dataframes

-   Suppose we have a second S&P 500 data located in a second Google Sheet and suppose that we would like to join or merge the data in this dataframe with the above dataframe sp500.

-   The ID of this second sheet is: `1F5KvFATcehrdJuGjYVqppNYC9hEKSww9rXYHCk2g6OA`

-   We can read the data present in this Google Sheet using the following code, similar to the one discussed above, using the following code.

```{r }
# Read additional S&P500 data that is posted in a Google Sheet.
library(gsheet)

prefix <- "https://docs.google.com/spreadsheets/d/"
sheetID <- "1nm688a3GsPM5cadJIwu6zj336WBaduglY9TSTUaM9jk"

# Form the URL to connect to
url <- paste(prefix, sheetID) 

# Read the Google Sheet located at the URL into a dataframe called gf
gf <- gsheet2tbl(url)
```

-   We now have two dataframes named `sp500` and `gf` that we wish to merge or join.

-   The two dataframes have a column named `Stock` in common, which will serve as the key, while doing the join.

-   The following code illusrates how to merge two dataframes:

```{r }
# merging dataframes
#df <- merge(sp500, gf , id = "Stock")
```

-   We now have a new dataframe named `df`, which contains the data got from merging the two dataframes `sp500` and `gf`. \[4\]

## Tibbles

1.  A tibble is a contemporary and enhanced variant of a R data frame that is part of the `tidyverse` package collection. \[5\]

2.  Tibbles are created and manipulated using the `dplyr` package, which provides a suite of functions optimized for data manipulation.

3.  The following characteristics distinguish a tibble from a conventional data frame:

4.  Tibbles must always have unique, non-empty column names. Tibbles do not permit the creation or modification of columns using partial matching of column names. Tibbles improve the output of large datasets by displaying by default only a few rows and columns.

5.  Tibbles have a more consistent behavior for subsetting.

6.  Here is an example of using the `tibble()` function from `dplyr` to construct a tibble:

```{r}
library(dplyr, warn.conflicts = FALSE)
# Create a tibble
my_tibble <- tibble(
  name = c("Ashok", "Bullu", "Charu"),
  age = c(72, 49, 46),
  gender = c("M", "F", "F")
)
# Print the tibble
my_tibble
```

-   This generates a tibble consisting of three columns (name, age, and gender) and three rows of data. Note that the column names are preserved and the tibble is printed in a compact and legible manner.

### Converting a dataframe into a tibble using `as_tibble()`

```{r}
# Create a data frame
my_df <- data.frame(
  name = c("Ashok", "Bullu", "Charu"),
  age = c(72, 49, 46),
  gender = c("M", "F", "F")
)
# Convert the data frame to a tibble
my_tibble <- as_tibble(my_df)
# Print the tibble
my_tibble
```

-   This assigns the tibble representation of the data frame `my_df` to the variable `my_tibble`.

-   Note that the resulting tibble has the same column names and data as the original data frame, but has the additional characteristics and behaviors of a tibble. \[5\]

### Converting a tibble into a dataframe

```{r}
library(dplyr)

# Convert the tibble to a data frame
my_df <- as.data.frame(my_tibble)

# Print the data frame
my_df
```

7.  A tibble offers several advantages over a data frame in R:

-   Large datasets can be printed with greater clarity and precision using Tibbles. By default, they only print the first few rows and columns, making it simpler to read and comprehend the data structure.

-   Better subsetting behavior: Tibbles have a more consistent subsetting behavior. This facilitates the subset and manipulation of data without unintended consequences.

-   Consistent naming: Tibbles always have column names that are distinct and non-empty. This makes it simpler to refer to specific columns and prevents errors caused by duplicate or unnamed column names.

-   More informative errors: Tibbles provides more informative error messages that make it simpler to diagnose and resolve data-related problems. \[5\]

## Summary of Chapter 4 -- Reading Data into R

In Chapter 4, we continue our exploration of data frames in R, focusing on reading various file formats into a data frame, managing the working directory, merging data frames, and the concept of tibbles, all with the overall objective of effectively reading data into R, for further analysis.

R uses the concept of a working directory to manage files. By default, R looks for and saves files in the working directory. To check the current working directory, we use the `getwd()` function. We can change the working directory using the `setwd()` function. It's advisable to choose an easily accessible working directory and avoid using spaces, special characters, and non-ASCII characters in file paths.

We can read various file formats into a data frame. For a CSV file, we use the `read.csv()` function. Reading an Excel file into a data frame is accomplished using the `read_excel` function from the `readxl` package. We also explored reading data from Google Sheets using the `gsheet2tbl` function in the `gsheet` package.

When dealing with multiple data sources, we may need to merge or join two data frames. R provides the `merge()` function to merge data frames based on a common key.

The final concept introduced in this chapter is tibbles. A tibble is a modern reimagining of the data frame, part of the tidyverse collection of packages. Tibbles are created and manipulated using the `dplyr` package. Tibbles have several distinct characteristics, such as unique, non-empty column names, better subsetting behavior, and improved output for large datasets. We reviewed how to create a tibble using the `tibble()` function, and convert a data frame to a tibble using the `as_tibble()` function, and vice versa with `as.data.frame()`.

In conclusion, this chapter built upon the concept of data frames and tibbles, showing how they can be used as a versatile tool for reading and storing data in various formats and getting ready to further explore the data.

## References

\[1\] RDocumentation. (n.d.). data.frame function. Retrieved from

[https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/data.frame"](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/data.frame%22)

"Tutorialspoint. (n.d.). R - Data Frames. Retrieved from <https://www.tutorialspoint.com/r/r_data_frames.htm>

Statistics Globe. (n.d.). What is a Data Frame in R? (3 Examples) \| data.frame Object Explained. Retrieved from <https://statisticsglobe.com/what-is-data-frame-r>

\[2\] McNeil, D. R. (1977). Interactive Data Analysis. Wiley. In *women: Average Heights and Weights for American Women*. Retrieved from <https://rdocumentation.org/packages/datasets/versions/3.6.2/topics/women>

Becker, R. A., Chambers, J. M., & Wilks, A. R. (1988). *The New S Language*. Wadsworth & Brooks/Cole. In *iris: Edgar Anderson's Iris Data*. Retrieved from <https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris>

Krasser, R. (2023, October 11). *Explore mtcars*. The Comprehensive R Archive Network. Retrieved from <https://cran.r-project.org/web/packages/explore/vignettes/explore_mtcars.html>

RDocumentation. (n.d.). *mtcars: Motor Trend Car Road Tests*. Retrieved from <https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars>

Henderson, V., & Velleman, P. (1981). *Motor Trend Car Road Tests*. Retrieved from <https://web.mit.edu/r/current/lib/R/library/datasets/html/mtcars.html>

Wickham, H. (2016). Ggplot2: Elegant graphics for data analysis (2nd ed.). Springer International Publishing.

\[3\] PsyTeachR. (n.d.). Chapter 1: Importing data from different file formats into R. Retrieved from <https://psyteachr.github.io>

Boston University School of Public Health. (n.d.). Reading and Writing Data to and from R. Retrieved from <http://sphweb.bumc.bu.edu>

RStudio. (n.d.). An Introduction to R - 7 Reading data from files. Retrieved from <https://rstudio.github.io>

Andrews, M. J. (n.d.). Read Multiple Files into a Single Data Frame. Retrieved from <http://www.mjandrews.org>

ProgrammingR. (n.d.). A Guide To File Readers And Data Conversion In R. Retrieved from <http://www.programmingr.com>

FreeCodeCamp. (n.d.). How to Work With Data Frames and CSV Files in R --- A Detailed Introduction with Examples. Retrieved from <http://www.freecodecamp.org>

R Core Team. (2021). getwd(): working directory; setwd(dir): change working directory. In R: A language and environment for statistical computing. R Foundation for Statistical Computing. Retrieved from https://stat.ethz.ch/R-manual/R-devel/library/base/html/getwd.html

\[4\] gsheet2tbl. (n.d.). In gsheet: Download Google Sheets Using Just the URL. Retrieved from <https://rdrr.io/cran/gsheet/man/gsheet2tbl.html>.

\[5\] Tibble overview and usage. (n.d.). In tibble: Simple Data Frames. Retrieved from [tibble.tidyverse.org](https://tibble.tidyverse.org/).

Analytics Steps. (n.d.). Tibbles in R Programming. Retrieved from [www.analyticssteps.com](https://www.analyticssteps.com/).

Tibble creation and name repair. (n.d.). In tibble: Build a data frame. Retrieved from [tibble.tidyverse.org](https://tibble.tidyverse.org/).

CRAN. (n.d.). Tibbles vs data frames: Printing and subsetting. Retrieved from [cran.r-project.org](https://cran.r-project.org/).
