# Live Case: S&P500 (3 of 3)

*Aug 10, 2023. -=-*

Agenda: Analyzing a particular Sector within the S&P500 Index *We have chosen to deeply analyze the HEALTH TECHNOLOGY Sector.*

## S&P 500 Data - PRELIMINARY SETUP

1.  We will continue our analysis of the S&P 500. Load the data, as described in the chapter Live Case: S&P500 (1 of 3)

```{r, warning=FALSE}
# Read S&P500 stock data present in a Google Sheet.
library(gsheet)
prefix <- "https://docs.google.com/spreadsheets/d/"
sheetID <- "11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM"
url500 <- paste(prefix,sheetID) # Form the URL to connect to
sp500 <- gsheet2tbl(url500) # Read it into a tibble called sp500
```

2.  Rename columns, as described in the chapter Live Case: S&P500 (1 of 3).

```{r, warning=FALSE}
suppressPackageStartupMessages(library(dplyr))

# Define a mapping of new column names
new_names <- c(
  "Date", "Stock", "StockName", "Sector", "Industry", 
  "MarketCap", "Price", "Low52Wk", "High52Wk", 
  "ROE", "ROA", "ROIC", "GrossMargin", 
  "OperatingMargin", "NetMargin", "PE", 
  "PB", "EVEBITDA", "EBITDA", "EPS", 
  "EBITDA_YOY", "EBITDA_QYOY", "EPS_YOY", 
  "EPS_QYOY", "PFCF", "FCF", 
  "FCF_QYOY", "DebtToEquity", "CurrentRatio", 
  "QuickRatio", "DividendYield", 
  "DividendsPerShare_YOY", "PS", 
  "Revenue_YOY", "Revenue_QYOY", "Rating"
)
# Rename the columns using the new_names vector
sp500 <- sp500 %>% 
  rename_with(~ new_names, everything())
```

```{r, warning=FALSE, eval=TRUE, echo=FALSE}
library(tibble)
new_names <- c(
  "Date", "Stock", "StockName", "Sector", "Industry", 
  "MarketCap", "Price", "Low52Wk", "High52Wk", 
  "ROE", "ROA", "ROIC", "GrossMargin", 
  "OperatingMargin", "NetMargin", "PE", 
  "PB", "EVEBITDA", "EBITDA", "EPS", 
  "EBITDA_YOY", "EBITDA_QYOY", "EPS_YOY", 
  "EPS_QYOY", "PFCF", "FCF", 
  "FCF_QYOY", "DebtToEquity", "CurrentRatio", 
  "QuickRatio", "DividendYield", 
  "DividendsPerShare_YOY", "PS", 
  "Revenue_YOY", "Revenue_QYOY", "Rating"
)
descriptions <- c(
  "Date (e.g. \"7/15/2023\")",
  "Stock Ticker (e.g. AAL)",
  "Name of the company (e.g \"American Airlines Group, Inc.\")",
  "Sector the stock belongs to (e.g. \"Transportation\")",
  "Industry the stock belongs to (e.g \"Airlines\")",
  "Market capitalization of the company",
  "Recent Stock Price",
  "52-Week Low Price",
  "52-Week High Price",
  "Return on Equity",
  "Return on Assets",
  "Return on Invested Capital",
  "Gross Profit Margin",
  "Operating Profit Margin",
  "Net Profit Margin",
  "Price-to-Earnings Ratio",
  "Price-to-Book Ratio",
  "Enterprise Value to EBITDA Ratio",
  "EBITDA",
  "Earnings per Share",
  "EBITDA Year-over-Year Growth",
  "EBITDA Quarterly Year-over-Year Growth",
  "EPS Year-over-Year Growth",
  "EPS Quarterly Year-over-Year Growth",
  "Price-to-Free Cash Flow",
  "Free Cash Flow",
  "Free Cash Flow Quarterly Year-over-Year Growth",
  "Debt-to-Equity Ratio",
  "Current Ratio",
  "Quick Ratio",
  "Dividend Yield",
  "Annual Dividends per Share Year-over-Year Growth",
  "Price-to-Sales Ratio",
  "Revenue Year-over-Year Growth",
  "Revenue Quarterly Year-over-Year Growth",
  "Technical Rating"
)
column_info <- tibble(ColumnName = new_names, Description = descriptions)
```

3.  Remove Rows containing no data or Null values, as described in the chapter Live Case: S&P500 (1 of 3).

```{r, warning=FALSE, eval=TRUE, echo=TRUE}
# Check for blank or null values in the "Stock" column
hasNull <- any(sp500$Stock == "" | is.null(sp500$Stock))
if (hasNull) { 
    # Remove rows with null or blank values from the dataframe tibble
    sp500 <- sp500[!(is.null(sp500$Stock) | sp500$Stock == ""), ]
}
```

4.  The S&P500 shares are divided into multiple Sectors. Thus, model Sector as a factor() variable, as described in the chapter Live Case: S&P500 (1 of 3).

```{r, warning=FALSE, eval=TRUE, echo=TRUE}
sp500$Sector <- as.factor(sp500$Sector)
```

5.  Stock Ratings: The S&P500 shares have Technical Ratings such as {Buy, Sell, ..}. Model the data column Rating as a factor() variable, as described in the chapter Live Case: S&P500 (1 of 3).

```{r, warning=FALSE, eval=TRUE, echo=TRUE}
sp500$Rating <- as.factor(sp500$Rating)
```

6.  Low52WkPerc: Create a new column to track Share Prices relative to their 52 Week Low, as described in the chapter Live Case: S&P500 (1 of 3).

```{r  warning=FALSE, eval=TRUE, echo=TRUE}
sp500 <- sp500 %>% mutate(Low52WkPerc = round((Price - Low52Wk)*100 / Low52Wk,2))
colnames(sp500)
```

Well done! Our data is now ready for analysis!!

7.  Creating a new column MarketCapBillions = MarketCap/1000,000,000

```{r}
sp500 <- sp500 %>% mutate(MarketCapBillions = MarketCap/ 1000000000)
colnames(sp500)

```

## SECTOR LEVEL ANALYSIS begins here

### Filter the data by sector Health Services, and display the number of stocks in the sector

```{r warning=FALSE, eval=TRUE, echo=TRUE}
ts <- sp500 %>%
		filter(Sector=='Health Services')

nrow(ts)
```

There are 12 number of of stocks in the sector Health Services

### Select the Specific Coulumns from the filtered dataframe `ts` (Health Services)

```{r warning=FALSE, eval=TRUE, echo=TRUE}
ts2 <- ts %>%
		select(Date, Stock, StockName,Sector,  Industry, MarketCap, Price,Low52Wk, High52Wk,
		       ROE, ROA,ROIC,GrossMargin, GrossMargin, 
		       NetMargin, Rating)

colnames(ts2)
```

### Arrange the Dataframe by ROE

```{r warning=FALSE, eval=TRUE, echo=TRUE}
ts3 <- ts2 %>% arrange(desc(ROE))
```

### Top 10 Shares in Sector Health Services Based on ROE

```{r warning=FALSE, eval=TRUE, echo=TRUE}
head(ts3,10)
```

### Mutate a data column called (Low52WkPerc), then show top 10 ROE stocks

```{r  warning=FALSE, eval=TRUE, echo=TRUE}
ts4 <- ts3 %>% mutate(Low52WkPerc = round((Price - Low52Wk)*100 / Low52Wk,2))
head(ts4[,c(1:3,10,16)],10)
```

### Summary Statistics of ROE

```{r}
ts3 <- na.omit(ts3)

ROESum <- ts3 %>%
  summarise(
    Mean = mean(ROE),
    Median= sd(ROE),
    Median= median(ROE),
    Q1 = quantile(ROE, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROE, probs = 0.75, na.rm = TRUE),
    Min = min(ROE),
    max = max(ROE)
  )

ROESum <- round(ROESum,2)
ROESum
```

### Summary Statistics of All key variables in Sector Health Services

```{r}
ts3 <- na.omit(ts3)

ROESum <- ts3 %>%
  summarise(
    Mean = mean(ROE),
    Median= sd(ROE),
    Median= median(ROE),
    Q1 = quantile(ROE, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROE, probs = 0.75, na.rm = TRUE),
    Min = min(ROE),
    max = max(ROE)
  )

ROESum <- round(ROESum,2)


ROASum <- ts3 %>%
  summarise(
    Mean = mean(ROA),
    Median= sd(ROA),
    Median= median(ROA),
    Q1 = quantile(ROA, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROA, probs = 0.75, na.rm = TRUE),
    Min = min(ROA),
    max = max(ROA)
  )

ROASum <- round(ROASum,2)

ROICSum <- ts3 %>%
  summarise(
    Mean = mean(ROIC),
    Median= sd(ROIC),
    Median= median(ROIC),
    Q1 = quantile(ROIC, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROIC, probs = 0.75, na.rm = TRUE),
    Min = min(ROIC),
    max = max(ROIC)
  )

ROICSum <- round(ROICSum,2)

GrossMarginSum <- ts3 %>%
  summarise(
    Mean = mean(GrossMargin),
    Median= sd(GrossMargin),
    Median= median(GrossMargin),
    Q1 = quantile(GrossMargin, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(GrossMargin, probs = 0.75, na.rm = TRUE),
    Min = min(GrossMargin),
    max = max(GrossMargin)
  )

GrossMarginSum <- round(GrossMarginSum,2)

NetMarginSum <- ts3 %>%
  summarise(
    Mean = mean(NetMargin),
    Median= sd(NetMargin),
    Median= median(NetMargin),
    Q1 = quantile(NetMargin, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(NetMargin, probs = 0.75, na.rm = TRUE),
    Min = min(NetMargin),
    max = max(NetMargin)
  )

NetMarginSum <- round(NetMarginSum,2)

Metrics <- c("ROE","ROA","ROIC","GrossMargin","NetMargin")

ftab <- rbind(ROESum, ROASum, ROICSum, GrossMarginSum, NetMarginSum)
ftab <- cbind(Metrics, ftab)
ftab
```

### Summary Statistics of ROE by each Sector of S&P500

```{r}

SectorROE <- sp500 %>%
  group_by(Sector) %>%
  summarise(
    Mean = mean(na.omit(ROE)),
    Median= sd(na.omit(ROE)),
    Median= median(na.omit(ROE)),
    Q1 = quantile(na.omit(ROE), probs = 0.25, na.rm = TRUE),
    Q3 = quantile(na.omit(ROE), probs = 0.75, na.rm = TRUE),
    Min = min(na.omit(ROE)),
    max = max(na.omit(ROE))
  )

cbind(Sector = SectorROE$Sector, round(SectorROE[,2:7],2))
```

## ANALYSIS OF HEALTH SERVICES SECTOR

1.  Market Cap of all companies in Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
library(janitor)
library(kableExtra)
# Market Cap by Stock
MCap <- ts3 %>%
  group_by(Stock) %>%
  summarise(
    MarketCapCr = sum(na.omit(MarketCap)/10000000))

# Sp500 Market Cap

SP500MarketCap <- sum(ts3$MarketCap/10000000)

# calculating % market cap
PercentMarketCap <- round(MCap$MarketCapCr*100/SP500MarketCap,2)
MCapTab <- cbind(MCap,PercentMarketCap)

# sorting by PercentMarketCap
MCapTab <- MCapTab %>% arrange(desc(PercentMarketCap))


MCapTab <- MCapTab %>%
  adorn_totals("row")

MCapTab <- knitr::kable(MCapTab, "html") %>% kable_styling() 
MCapTab 
```

2.  Shares which are most attractively priced in Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(Low52WkPerc)
AttrShares <- AttrShares[, c(2:4,7,8,10,11,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```

### PROFITABILITY OF HEALTH SERVICES SECTOR

1.  Shares have highest ROE within Sector Technology Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(desc(ROE))
AttrShares <- AttrShares[, c(2:4,7,8,10,11,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```

2.  Shares have highest ROA within Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(desc(ROA))
AttrShares <- AttrShares[, c(2:4,7,8,10,11,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```

3.  Shares have highest NetMargin within Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(desc(NetMargin))
AttrShares <- AttrShares[, c(2:4,7,8,10,11,14,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```
