# Live Case: S&P500 (REITs)

*Sep 05, 2023*

## Objective


### A1) Role Play?

Greetings Data Commandos!

Imagine you're in the bustling hub of the world's most elite consulting firms, revered across the corporate spectrum.

**A prestigious investment fund is ready to channel $1 Million into the US finance sector. They've enlisted your expertise to delve into the 29 REITs within the Finance sector of the S&P500.**  

Real Estate Investment Trusts, commonly known as REITs, offer a distinctive way to engage with real estate markets without the cumbersome process of directly owning property.

**Your mission? Allocate `$1 Million` to the "best" REIT(s). Pinpoint the top 1, 2 or 3 REITs that present the most promising short-term trading opportunities. Dive in and make those data-driven decisions!**


### A2) What are Prof. Sameer's learning objectives for you?

1. Revisit and solidify your knowledge of R programming, acquired or expected to be acquired in earlier courses.
2. Enhance your proficiency in data management and manipulation using the `dplyr` package and other functions in R.
3. Sharpen your skills in data visualization leveraging the `ggplot2`, `ggpubr` and related packages in R.
4. Master the art of addressing a data-centric business challenge while embodying the role of a Consulting Team.
5. Cultivate the capability to compellingly present your solutions to a discerning yet just audience.





## Setup S&P 500 **REIT** Data 

### 1. Load some useful R packages
```{r, warning=FALSE}
# Load the required libraries, suppressing annoying startup messages
library(dplyr, quietly = TRUE, warn.conflicts = FALSE) # For data manipulation
library(tibble, quietly = TRUE, warn.conflicts = FALSE) # For data manipulation
library(ggplot2, quietly = TRUE, warn.conflicts = FALSE) # For data visualization
library(ggpubr, quietly = TRUE, warn.conflicts = FALSE) # For data visualization

library(gsheet, quietly = TRUE, warn.conflicts = FALSE) # For Google Sheets
library(rmarkdown, quietly = TRUE, warn.conflicts = FALSE) # For writing
library(knitr, quietly = TRUE, warn.conflicts = FALSE) # For tables
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE) # For tables
```

### 2. Read S&P500 data (to derive REIT data)
```{r, warning=FALSE}
# Read S&P500 stock data present in a Google Sheet.
library(gsheet)
prefix <- "https://docs.google.com/spreadsheets/d/"
sheetID <- "11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM"
url500 <- paste(prefix,sheetID) # Form the URL to connect to
sp500 <- gsheet2tbl(url500) # Read it into a tibble called sp500
```

### 3. Rename Columns
```{r, warning=FALSE}
# Define a mapping of new column names
new_names <- c(
  "Date", "Stock", "StockName", "Sector", "Industry", 
  "MarketCap", "Price", "Low52Wk", "High52Wk", 
  "ROE", "ROA", "ROIC", "GrossMargin", 
  "OperatingMargin", "NetMargin", "PE", 
  "PB", "EVEBITDA", "EBITDA", "EPS", 
  "EBITDA_YOY", "EBITDA_QYOY", "EPS_YOY", 
  "EPS_QYOY", "PFCF", "FCF", 
  "FCF_QYOY", "DebtToEquity", "CurrentRatio", 
  "QuickRatio", "DividendYield", 
  "DividendsPerShare_YOY", "PS", 
  "Revenue_YOY", "Revenue_QYOY", "Rating"
)
# Rename the columns using the new_names vector
colnames(sp500)<-new_names
```



### 3. Select REIT data



1. The S&P500 shares are divided into multiple Sectors. Each stock belongs to a unique sector. Thus, it makes sense to model Sector as a factor() variable.

1. Set Sector, Rating data columns to be factor data types.

```{r}
sp500$Sector <- as.factor(sp500$Sector)
sp500$Rating <- as.factor(sp500$Rating)
```


### C6. The **Finance** Sector within the S&P500 

1. The Finance sector plays a pivotal role in the overall U.S. economy. Its performance is often closely watched by economists and investors alike, given its profound impact on lending, investment, and overall economic growth. Over the years, regulatory changes, monetary policy, and global economic events have significantly influenced this sector, making it a dynamic and critical component of the S&P 500.

2. We focus on investment opportunities **within the Finance sector** of the S&P500.

* We want to determine the *fundamentally strongest* AND *most reasonably priced* shares for *short-to-medium term* investing.

3. **Industry**:  The Finance sector includes many industries within it. 

For example:

   - **Banks**: JPMorgan Chase, Bank of America, and Wells Fargo, among others, represent the significant banking entities. 
   
   - **Insurance Companies**: Companies like Berkshire Hathaway, Allstate operate in this sub-sector, offering a range of insurance products from property and casualty insurance to life insurance.
   
4. We create a tibble named `finStocks`, filtering the shares that belong to the Finance sector.

```{r}
finStocks = sp500 %>% 
  filter(Sector=="Finance") 
```

5. Industries within the Finance Sector

* The data shows that the Finance sector consists of a total of `r nrow(finStocks)` companies that belong to different Industries. 
* We set the Industry to be factor variable, since they can only assume unique levels.

```{r}
finStocks$Industry <- as.factor(finStocks$Industry)
```

* The following visualization summarizes the different Industries within the Finance Sector:

```{r}
ggplot(data = finStocks, 
       aes(y = Industry)) + 
  geom_bar(aes(fill = Industry)) +
  geom_text(stat='count', 
            aes(label=after_stat(count))) +
  labs(title = "Stocks by Industry, within Finance Sector", 
       x = "No. of Stocks", 
       y = "Industry")
```


* Market Capitalization (Billions of USD) of Industries within the Finance Sector

```{r}
FinanceMarketCap <- finStocks %>%
  mutate(MarketCap_Billions = round(MarketCap/1000000000, 2)) %>%
  group_by(Industry) %>%
  summarise(Market_Cap_BillionUSD = sum(MarketCap_Billions, na.rm = TRUE))  %>%
  arrange(-Market_Cap_BillionUSD)  

# Create a summary row
summary_row <- tibble(
  Industry = "Total",
  Market_Cap_BillionUSD = sum(FinanceMarketCap$Market_Cap_BillionUSD)
)
# Append the summary row to the result
FinanceMarketCap <- bind_rows(FinanceMarketCap, summary_row)

# Render the table
FinanceMarketCap %>%
  kable("html", caption = "Market Capitalization (Billions of USD) of Finance Sector") %>% 
  kable_styling()
```


* We focus on investment opportunities within a particular Industry -- **Real Estate Investment Trusts**.


### C7. Stock Prices, as of `r finStocks$Date[1]`

1. Stock Prices relative to their 52 Week Low and 52 Week High

* We want to analyze stock prices relative to their 52 Week Low and 52 Week High respectively, to understand their relative price attractiveness.

* For this purpose, we create some additional data columns.

```{r}
finStocks = sp500 %>% 
  filter(Sector=="Finance") %>% 
  mutate(Low52WkPerc = round((Price - Low52Wk)*100 / Low52Wk,2)) %>% 
  mutate(High52WkPerc = round((High52Wk - Price)*100 / Low52Wk,2)) %>%
  mutate(MarketCap_Billions = round(MarketCap/1000000000, 2))
```


* Here, a new column named `Low52WkPerc` is being added. The column contains the percentage change between the current price (`Price`) and its 52-week low (`Low52Wk`). The formula used is: $$Low52WkPerc = \frac{(CurrentPrice - 52WeekLow)*100}{52WeekLow}$$

* Another column named `High52WkPerc` represents the percentage change between the 52-week high (`High52Wk`) and the current price (`Price`). 

* We round off the data to two decimal places for clarity. 

```{r}
finStocks$Price <- round(finStocks$Price,1)
finStocks$Low52Wk <- round(finStocks$Low52Wk,1)
finStocks$High52Wk <- round(finStocks$High52Wk,1)
```






\newpage

## D. REITs in the S&P500, as of `r finStocks$Date[1]`

1. Real Estate Investment Trusts, commonly known as REITs, stand as a cornerstone for investors seeking diversification in their portfolios. These entities offer a distinctive way to engage with real estate markets without the cumbersome process of directly owning property.

2. In our analysis of the Finance sector, we want to focus attention on a particular Industry within it -- **Real Estate Investment Trusts**.

3. Recall: We want to determine the *fundamentally strongest* AND *most reasonably priced*, top 1-3 REITs for *short-to-medium term* investing USD 1 Million.

### D1. **Key Characteristics of REITs:**

1. **Income Distribution**: One of the most touted features of REITs is their consistent income flow. U.S. tax regulations mandate REITs to distribute at least 90% of their taxable income as dividends. While this can be enticing due to potentially higher yields, it also poses a risk. The high dividend mandate leaves REITs with less retained earnings, potentially hindering their growth or making them more dependent on external financing.

2. **Liquidity versus Direct Ownership**: REITs offer a stark contrast to traditional real estate investments in terms of liquidity. While selling a property might entail prolonged durations, hefty transaction costs, and price negotiations, REIT shares can be traded with the agility of stocks. This flexibility, however, comes at the cost of exposure to stock market volatility.

3. **Tax Implications**: The unique tax structure of REITs is a double-edged sword. While they can dodge corporate taxes by abiding by stringent regulations, such as the income distribution clause, shareholders often have to pay higher individual taxes on REIT dividends compared to qualified stock dividends.

4. **Sectoral Diversification**: REITs don't just represent traditional brick-and-mortar assets. From data centers to timberlands, they span diverse sectors, potentially providing portfolio diversification. However, the granularity in sectors necessitates that investors be judicious and knowledgeable about the specific type of real estate exposure they're obtaining.

### D2. **Major U.S. REITs:**

1. **American Tower Corporation (AMT)**: Pioneering the realm of communication infrastructures, AMT emphasizes cell tower operations. While it highlights the evolution of REITs beyond traditional confines, it also underscores the need for REIT investors to comprehend tech industry dynamics, given its tech infrastructure focus.

2. **Prologis (PLD)**: With a niche in logistics and industrial real estate, Prologis stands out in the age of e-commerce. The company's assets, mainly distribution centers, are strategically situated in prime markets. However, the increasing demand for same-day deliveries and supply chain revamps could challenge Prologis' portfolio.

3. **Simon Property Group (SPG)**: Catering predominantly to retail spaces, SPG faces the arduous task of reinventing malls in an era where brick-and-mortar stores battle online retailers. The company's resilience in nurturing mixed-use spaces might determine its long-term growth trajectory.

4. **Equity Residential (EQR)**: As urbanization continues, EQR's focus on high-density urban areas might seem lucrative. But, with telecommuting trends and urban exodus, it's pivotal to monitor how urban rental landscapes evolve.

5. **Digital Realty Trust (DLR)**: In the digital age, DLR taps into the data economy by majoring in data centers. While the tech boom supports such endeavors, DLR's growth could be contingent on global data regulations and tech infrastructure demands.

**References**: Please consider looking into the following well-known sources, which regularly publish information about REITs. 

* National Association of Real Estate Investment Trusts (NAREIT) -- This organization is a representative voice for REITs in the U.S. They frequently release reports, articles, and data on the REIT industry.

* Major Financial News Outlets - Outlets like The Wall Street Journal, Financial Times, and Bloomberg often feature articles on REITs, especially in their real estate or investment sections.

* The Journal of Real Estate Finance and Economics - This academic journal covers a wide range of topics in real estate, including REITs. 


### D3. **REITs in the S&P500:**

* We create a tibble named `REIT` from within the Finance sector tibble `finStocks`. Specifically, we filter the shares that belong to the `Real Estate Investment Trusts` Industry, within the Finance sector.

```{r}
REIT <- finStocks %>% 
  filter(Industry == 'Real Estate Investment Trusts')
```

* The following table lists REITs within the Finance sector of the S&P500

```{r}
REIT %>% 
  select(Stock, StockName, Price, MarketCap_Billions)  %>%
  arrange(desc(MarketCap_Billions)) %>%
  kable("html", caption = "REITs within Finance Sector of S&P500") %>% 
  kable_styling()
```

* Consider the summary statistics of the Market Capitalization of the REITs within the S&P500

```{r echo=TRUE}
REIT %>% summarise(
  N = n(),
  Mean = mean(MarketCap_Billions),
  SD = sd(MarketCap_Billions),
  Median = median(MarketCap_Billions),
  Q1 = quantile(MarketCap_Billions, 0.25),
  Q3 = quantile(MarketCap_Billions, 0.75),
  Min = min(MarketCap_Billions),
  Max = max(MarketCap_Billions),
  Sum = sum(MarketCap_Billions)
) %>% 
  round(2) %>%
  kable("html", caption = "Summary Statistics of Market Capitalizaiton of REITs (Billion USD)") %>% 
  kable_styling()
```
* As can be seen, the S&P500 consists of `r nrow(REIT)` REITs. 

**Recall.. A prestigious investment fund is ready to channel $1 Million into the US finance sector. They've enlisted your expertise to delve into the 29 REITs within the Finance sector of the S&P500.**  


**Recall Your mission? Pinpoint the top 1, 2 or 3 REITs that present the most promising short-term trading opportunities. How will you allocate your investment capital of USD 1 Million? Dive in and make those data-driven decisions!**

* We want to determine the *fundamentally strongest* AND *most reasonably priced* shares for *short-to-medium term* investing.



Well done! Our data is now ready for analysis!!


7.  Low52WkPerc: Create a new column MarketCapBillions = MarketCap/1000,000,000, as described in the chapter Live Case: S&P500 (1 of 3).

```{r  warning=FALSE, eval=TRUE, echo=TRUE}
sp500 <- sp500 %>% mutate(MarketCapBillions = round(MarketCap/1000000000))
colnames(sp500)
```

## Live Case: S&P500


ISSUE: Analysis of a particular SECTOR We have chosen to deeply analyze the HEALTH TECHNOLOGY Sector

## SECTOR LEVEL ANALYSIS begins here

### Filter the data by sector Health Services, and display the number of stocks in the sector

```{r warning=FALSE, eval=TRUE, echo=TRUE}
ts <- sp500 %>%
		filter(Sector=='Health Services')

nrow(ts)
```

There are 12 number of of stocks in the sector Health Services

### Select the Specific Coulumns from the filtered dataframe `ts` (Health Services)

```{r warning=FALSE, eval=TRUE, echo=TRUE}
ts2 <- ts %>%
		select(Date, Stock, StockName,Sector,  Industry, MarketCap, Price,Low52Wk, High52Wk,
		       ROE, ROA,ROIC,GrossMargin, GrossMargin, 
		       NetMargin, Rating)

colnames(ts2)
```

### Arrange the Dataframe by ROE

```{r warning=FALSE, eval=TRUE, echo=TRUE}
ts3 <- ts2 %>% arrange(desc(ROE))
```

### Top 10 Shares in Sector Health Services Based on ROE

```{r warning=FALSE, eval=TRUE, echo=TRUE}
head(ts3,10)
```

### Significance of 52-Week Low Price

The 52-week low price of a stock is a significant indicator for multiple reasons, especially when considering shares listed on major indices like the S&P 500. Here's why this metric is noteworthy:

1. **Historical Perspective**: The 52-week low offers a snapshot of how low the stock has traded over the past year relative to its current price, providing context about its price journey.

2. **Potential Entry Point**: Some investors view stocks that are near their 52-week low as potential buying opportunities, under the assumption that the stock might be undervalued and could rebound.

3. **Psychological Level**: Stocks approaching their 52-week low can be seen as testing a significant support level. If a stock consistently fails to breach its 52-week low, it might indicate that the market values the stock at that level, and it's resistant to falling below it.

4. **Basis for Technical Analysis**: For technical analysts or traders, the 52-week low serves as a critical reference point. A consistent breach of this level might signify a bearish trend, while a rebound can indicate potential recovery.

5. **Yield Implications for Dividend Stocks**: For dividend-paying stocks, a price near the 52-week low (assuming the dividend hasn't been cut) would imply a higher dividend yield, potentially making it attractive for income-seeking investors.

* **Note of Caution:** While the 52-week low is a valuable reference point, it's essential to interpret it in conjunction with other financial and market indicators. A stock trading near its 52-week low doesn't automatically make it a good buy, just as a stock trading near its 52-week high doesn't automatically make it overvalued. Comprehensive analysis, should inform investment decisions.


### Mutate a data column called (Low52WkPerc), then show top 10 ROE stocks


```{r  warning=FALSE, eval=TRUE, echo=TRUE}
ts4 <- ts3 %>% mutate(Low52WkPerc = round((Price - Low52Wk)*100 / Low52Wk,2))
head(ts4[,c(1:3,10,16)],10)
```
## Summary Statistics of Low52WkPerc (Price rel. to 52-Week Low)

```{r eval=TRUE, echo=TRUE, warning=FALSE}
summaryStats <- ts4 %>% summarise(
  N = n(),
  Mean = mean(Low52WkPerc),
  SD = sd(Low52WkPerc),
  Median = median(Low52WkPerc),
  Q1 = quantile(Low52WkPerc, 0.25),
  Q3 = quantile(Low52WkPerc, 0.75),
  Min = min(Low52WkPerc),
  Max = max(Low52WkPerc)
) 

Low52WkPercQ1 <- summaryStats$Q1 # Save Q1 of Low52WkPerc

summaryStats %>% 
  round(2) %>%
  kable("html", caption = "Summary Statistics of Low52WkPerc (Price rel. to 52-Week Low)") %>% 
  kable_styling()
```


*Low52WkPerc for all the Health Sector Stocks, as shown below*

```{r eval=TRUE, echo=FALSE, warning=FALSE}
PlotBox <- ggboxplot(ts4, 
            y = "Low52WkPerc", 
            rug = TRUE, 
            color = "black",
            fill = "gold",
            add = "jitter",
            orientation = "horizontal",
            title = "Box Plot showing (Median, Q1, Q3) of Low52WkPerc"
)
```

```{r  eval=TRUE, echo=FALSE, warning=FALSE}
# Create a summary data frame
data_summary <- ts4 %>%
  summarise(
    MeanLow52WkPerc = mean(Low52WkPerc, na.rm = TRUE),
    SDLow52WkPerc = sd(Low52WkPerc, na.rm = TRUE)
  )

PlotBar <- ggplot(data_summary, 
       aes(y = "", 
           x = MeanLow52WkPerc)) +
  geom_bar(stat = "identity", 
           fill = "lightpink", color = "lightpink") +
  geom_errorbar(aes(xmin = MeanLow52WkPerc - SDLow52WkPerc, 
                    xmax = MeanLow52WkPerc + SDLow52WkPerc), 
                width = .2) +
  labs(x = "Low52WkPerc", y = "",
       title = "Bar Plot showing (Mean +/- SD) of Low52WkPerc") +
  theme_minimal() 
```

```{r eval=TRUE, echo=FALSE, warning=FALSE}
# Combine the plots using ggarrange()
ggarrange(PlotBox, PlotBar, nrow = 2)
```

## Inexpensive Stocks with Low52WkPerc < Q1(Low52WkPerc)

```{r eval=TRUE, echo=TRUE, warning=FALSE}
ts4 %>% 
  select(Stock, StockName, Price, Low52Wk, Low52WkPerc) %>% 
  filter(Low52WkPerc < Low52WkPercQ1) %>% 
  arrange(Low52WkPerc)%>%
  kable("html", caption = "Inexpensive Stocks with Low52WkPerc < Q1(Low52WkPerc)") %>% 
  kable_styling()
```



### Summary Statistics of Return on Equity (ROE)

```{r}
summaryStats <- ts4 %>% summarise(
  N = n(),
  Mean = mean(ROE, na.rm = TRUE),
  SD = sd(ROE, na.rm = TRUE),
  Median = median(ROE, na.rm = TRUE),
  Q1 = quantile(ROE, 0.25, na.rm = TRUE),
  Q3 = quantile(ROE, 0.75, na.rm = TRUE),
  Min = min(ROE, na.rm = TRUE),
  Max = max(ROE, na.rm = TRUE)
) 

ROE_Q3 <- summaryStats$Q3

summaryStats %>% 
  round(2) %>%
  kable("html", caption = "Summary Statistics of Return on Equity (ROE)") %>% 
  kable_styling()
```

-   ROE for all the Stocks in Health Sector, as shown below

```{r eval=TRUE, echo=FALSE, warning=FALSE}
PlotBox <- ggboxplot(ts4, 
            y = "ROE", 
            rug = TRUE, 
            color = "black",
            fill = "gold",
            add = "jitter",
            orientation = "horizontal",
            title = "Box Plot showing (Median, Q1, Q3) of ROE"
)
```

```{r eval=TRUE, echo=FALSE, warning=FALSE}
# Create a summary data frame
data_summary <- ts4 %>%
  summarise(
    MeanROE = mean(ROE, na.rm = TRUE),
    SDROE = sd(ROE, na.rm = TRUE)
  )
# Generate a Bar Plot
PlotBar <- ggplot(data_summary, 
       aes(y = "", 
           x = MeanROE)) +
  geom_bar(stat = "identity", 
           fill = "lightpink", color = "lightpink") +
  geom_errorbar(aes(xmin = MeanROE - SDROE, 
                    xmax = MeanROE + SDROE), 
                width = .2) +
  labs(x = "ROE", y = "",
       title = "Bar Plot showing (Mean +/- SD) of ROE") +
  theme_minimal() 
```

```{r eval=TRUE, echo=FALSE, warning=FALSE}
# Combine the plots using ggarrange()
ggarrange(PlotBox, PlotBar, nrow = 2)
```

### Stocks with ROE > Q3(ROE)

```{r}
ts4 %>% 
  select(Stock, StockName, Price, ROA, ROE, Low52Wk, Low52WkPerc) %>% 
  filter(ROE > ROE_Q3) %>%
  arrange(desc(ROE)) %>%
  kable("html", caption = "Stocks with ROE > Q3(ROE)") %>% 
  kable_styling()

```


### Summary Statistics of Return on Equity (ROA)

```{r}
summaryStats <- ts4 %>% summarise(
  N = n(),
  Mean = mean(ROA, na.rm = TRUE),
  SD = sd(ROA, na.rm = TRUE),
  Median = median(ROA, na.rm = TRUE),
  Q1 = quantile(ROA, 0.25, na.rm = TRUE),
  Q3 = quantile(ROA, 0.75, na.rm = TRUE),
  Min = min(ROA, na.rm = TRUE),
  Max = max(ROA, na.rm = TRUE)
) 

ROA_Q3 <- summaryStats$Q3

summaryStats %>% 
  round(2) %>%
  kable("html", caption = "Summary Statistics of Return on Equity (ROA)") %>% 
  kable_styling()
```

-   ROA for all the Stocks in Health Sector, as shown below

```{r eval=TRUE, echo=FALSE, warning=FALSE}
PlotBox <- ggboxplot(ts4, 
            y = "ROA", 
            rug = TRUE, 
            color = "black",
            fill = "gold",
            add = "jitter",
            orientation = "horizontal",
            title = "Box Plot showing (Median, Q1, Q3) of ROA"
)
```

```{r eval=TRUE, echo=FALSE, warning=FALSE}
# Create a summary data frame
data_summary <- ts4 %>%
  summarise(
    MeanROA = mean(ROA, na.rm = TRUE),
    SDROA = sd(ROA, na.rm = TRUE)
  )
# Generate a Bar Plot
PlotBar <- ggplot(data_summary, 
       aes(y = "", 
           x = MeanROA)) +
  geom_bar(stat = "identity", 
           fill = "lightpink", color = "lightpink") +
  geom_errorbar(aes(xmin = MeanROA - SDROA, 
                    xmax = MeanROA + SDROA), 
                width = .2) +
  labs(x = "ROA", y = "",
       title = "Bar Plot showing (Mean +/- SD) of ROA") +
  theme_minimal() 
```

```{r eval=TRUE, echo=FALSE, warning=FALSE}
# Combine the plots using ggarrange()
ggarrange(PlotBox, PlotBar, nrow = 2)
```

### Stocks with ROA > Q3(ROA)

```{r}
ts4 %>% 
  select(Stock, StockName, Price, ROA, ROE, Low52Wk, Low52WkPerc) %>% 
  filter(ROA > ROA_Q3) %>%
  arrange(desc(ROA)) %>%
  kable("html", caption = "Stocks with ROA > Q3(ROA)") %>% 
  kable_styling()

```



### ROE versus ROA and colored by Price rel. to 52 Week Low

```{r, warning=FALSE}
top10 <- 
  ts4 %>% 
  select(Stock, Price, Low52Wk, Low52WkPerc, ROA, ROE) %>% 
  arrange(desc(ROE))%>%
  slice(1:10)

top10$name <- top10$Stock

ggscatter(top10, 
          x = "ROA", 
          y = "ROE", 
          size = "Low52WkPerc",
          color = "Low52WkPerc",
          alpha = 0.5,
          label = "name", 
          repel = TRUE,
          title = "ROE vs ROA, Low52WkPerc for Health Sector with highest ROE") + 
  gradient_color(c("darkgreen",  "red"))
```

### Summary Statistics of All key variables in Sector Health Services

```{r}
ts3 <- na.omit(ts3)

ROESum <- ts3 %>%
  summarise(
    Mean = mean(ROE),
    Median= sd(ROE),
    Median= median(ROE),
    Q1 = quantile(ROE, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROE, probs = 0.75, na.rm = TRUE),
    Min = min(ROE),
    max = max(ROE)
  )

ROESum <- round(ROESum,2)


ROASum <- ts3 %>%
  summarise(
    Mean = mean(ROA),
    Median= sd(ROA),
    Median= median(ROA),
    Q1 = quantile(ROA, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROA, probs = 0.75, na.rm = TRUE),
    Min = min(ROA),
    max = max(ROA)
  )

ROASum <- round(ROASum,2)

ROICSum <- ts3 %>%
  summarise(
    Mean = mean(ROIC),
    Median= sd(ROIC),
    Median= median(ROIC),
    Q1 = quantile(ROIC, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(ROIC, probs = 0.75, na.rm = TRUE),
    Min = min(ROIC),
    max = max(ROIC)
  )

ROICSum <- round(ROICSum,2)

GrossMarginSum <- ts3 %>%
  summarise(
    Mean = mean(GrossMargin),
    Median= sd(GrossMargin),
    Median= median(GrossMargin),
    Q1 = quantile(GrossMargin, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(GrossMargin, probs = 0.75, na.rm = TRUE),
    Min = min(GrossMargin),
    max = max(GrossMargin)
  )

GrossMarginSum <- round(GrossMarginSum,2)

NetMarginSum <- ts3 %>%
  summarise(
    Mean = mean(NetMargin),
    Median= sd(NetMargin),
    Median= median(NetMargin),
    Q1 = quantile(NetMargin, probs = 0.25, na.rm = TRUE),
    Q3 = quantile(NetMargin, probs = 0.75, na.rm = TRUE),
    Min = min(NetMargin),
    max = max(NetMargin)
  )

NetMarginSum <- round(NetMarginSum,2)

Metrics <- c("ROE","ROA","ROIC","GrossMargin","NetMargin")

ftab <- rbind(ROESum, ROASum, ROICSum, GrossMarginSum, NetMarginSum)
ftab <- cbind(Metrics, ftab)
ftab
```


## ANALYSIS OF HEALTH SERVICES SECTOR

1.  Market Cap of all companies in Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
library(janitor)
library(kableExtra)
# Market Cap by Stock
MCap <- ts3 %>%
  group_by(Stock) %>%
  summarise(
    MarketCapBi = round(sum(na.omit(MarketCap)/1000000000),2))

# Sp500 Market Cap

SP500MarketCap <- sum(ts3$MarketCap/1000000000)

# calculating % market cap
PercentMarketCap <- round(MCap$MarketCapBi*100/SP500MarketCap,2)
MCapTab <- cbind(MCap,PercentMarketCap)

# sorting by PercentMarketCap
MCapTab <- MCapTab %>% arrange(desc(PercentMarketCap))


MCapTab <- MCapTab %>%
  adorn_totals("row")

MCapTab <- knitr::kable(MCapTab, "html") %>% kable_styling() 
MCapTab 
```

2.  Shares which are most attractively priced in Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(Low52WkPerc)
AttrShares <- AttrShares[, c(2:4,7,8,10,11,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```

### PROFITABILITY OF HEALTH SERVICES SECTOR

1.  Shares have highest ROE within Sector Technology Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(desc(ROE))
AttrShares <- AttrShares[, c(2:4,7,8,10,11,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```

2.  Shares have highest ROA within Sector Health Services

```{r  warning=FALSE, eval=TRUE, echo=TRUE,message=FALSE}
AttrShares <- ts4 %>% arrange(desc(ROA))
AttrShares <- AttrShares[, c(2:4,7,8,10,11,16)]

AttrShares <- knitr::kable(AttrShares, "html") %>% kable_styling() 
AttrShares 
```